---
title: "Colours & plot styles"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 100,
  out.width = "100%"
)
```

This article covers the ways to colour a rainbow plot: the rotating gradient and its
period, custom ramps, the shipped colour-blind palettes, discrete **rank** colours (the
"boring" fixed-colour mode), and **shade** plots.

```{r setup}
library(HaplotypeRainbows)
library(ggplot2)

data("pfisolateExample")
rb <- HaplotypeRainbow$new(
  pfIsosHeomeV1,
  sample_col    = "s_Sample",
  target_col    = "p_name",
  popuid_col    = "h_popUID",
  rel_abund_col = "c_AveragedFrac"
)
rb$prep(sort = "population_rank")
```

## The rainbow gradient

By default the dominant haplotype's hue rotates across targets with a **period**
(`color_period`, default 11), which is what produces the "rainbow". A shorter period
cycles the colours faster:

```{r color-period}
rb$prep(sort = "population_rank", color_period = 6)
rb$plot(x_axis_labels = FALSE)
```

```{r color-period-reset, include = FALSE}
rb$prep(sort = "population_rank")
```

Supply your own ramp with `colors` (its length should match `color_period`):

```{r colors-custom}
rb$plot(colors = c("#F50300", "#FF6E00", "#FFEB01", "#00CA1E", "#0241FE", "#FE00D4"))
```

The fill can be driven by either colour column from the prep — log-spaced (default) or
linear hue:

```{r color-col, eval = FALSE}
rb$plot(color_col = "popidFracLogColor")   # default (log-spaced hue)
rb$plot(color_col = "popidFracRegColor")   # linear hue
```

## Colour-blind-friendly palettes

The package ships three categorical palettes (`colorPalette_08`, `colorPalette_12`,
`colorPalette_15`) used by the metadata sidebars and available for your own use:

```{r palettes}
scales::show_col(colorPalette_12)
```

## Discrete rank colours (the "boring" plot)

`plot(rank_colors = TRUE)` colours haplotypes by a discrete **rank** with a legend, so the
most abundant haplotype (rank 1) is the *same* colour in every target — a flat,
easy-to-read alternative to the gradient.

When there are more ranks than palette colours, a ramp is interpolated from the palette
(dropping white/black, ordering by hue, then reordering so consecutive ranks stay
distinct — the abundant ranks 1, 2, 3 won't come out near-identical). Prep with
`mark_invariant = TRUE` to give single-haplotype (invariant) targets their own colour
(black by default).

```{r rank-colors, fig.width = 14, fig.height = 9}
rb$prep(sort = "population_rank", mark_invariant = TRUE)$sort_by_clustering()
rb$plot(rank_colors = TRUE, x_axis_labels = FALSE)
```

`get_rank_colors()` returns the rank -> colour map (e.g. to reuse elsewhere); pass a
different `rank_palette` to `plot()` / `get_rank_colors()` to change it:

```{r rank-map}
head(rb$get_rank_colors(), 5)
```

## Shade plots

Instead of a rotating rainbow, `prep_shade()` colours each target with **shades** of a
per-target base colour. `plot()` then uses the shade colours automatically:

```{r shade}
rb$prep_shade(min_pop_size = 1)
rb$plot(x_axis_labels = FALSE)
```

Pass your own base colours (each target cycles through these, then shades within):

```{r shade-base}
rb$prep_shade(base_colors = colorPalette_08)
rb$plot(x_axis_labels = FALSE)
```
