---
title: "Examples: a full tour of the options"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 100,
  out.width = "100%"
)
```

This article is the exhaustive companion to the README: it walks through **every**
`HaplotypeRainbow` option so the README can stay slim. Each section is a self-contained
recipe you can copy.

```{r setup}
library(HaplotypeRainbows)
library(ggplot2)

# example data (uses the older SeekDeep column names)
data("pfisolateExample")
```

## Constructing the object

`HaplotypeRainbow` carries your column mapping so you only set it once. The defaults
follow the Portable Microhaplotype Object (PMO) convention
(`library_sample_name` / `target_name` / `seq` / `reads`); the example data uses the
older SeekDeep names, so we pass them explicitly.

```{r construct}
rb <- HaplotypeRainbow$new(
  pfIsosHeomeV1,
  sample_col    = "s_Sample",
  target_col    = "p_name",
  popuid_col    = "h_popUID",
  rel_abund_col = "c_AveragedFrac"
)
rb
```

`haplotype_rainbow()` is an identical convenience constructor:

```{r construct-wrapper, eval = FALSE}
rb <- haplotype_rainbow(my_pmo_allele_table)   # PMO column names, no mapping needed
```

Transforming methods (`prep()`, `prep_shade()`, `sort_*()`, `set_sample_order()`) mutate
the object and return it, so they chain; `plot()` returns a ggplot.

## Prepping: haplotype ordering

`prep(sort = ...)` controls how haplotypes stack within each cell.

```{r prep-population}
rb$prep(sort = "population_rank")   # order by population rank (default)
rb$plot()
```

```{r prep-frac}
rb$prep(sort = "within_sample_freq")   # order by within-sample fraction
rb$plot()
```

Other `prep()` knobs:

```{r prep-args, eval = FALSE}
rb$prep(
  sort         = "population_rank",
  min_pop_size = 2,     # drop targets with < 2 unique haplotypes
  color_period = 11,    # hue steps to rotate across targets (the "rainbow" period)
  bar_height   = 0.80   # < 1 leaves a gap between sample rows
)
```

## Sorting samples

By default samples are alphabetical. Cluster similar samples (by haplotype sharing,
ward.D2) so they sit together:

```{r sort-cluster}
rb$prep(sort = "population_rank")$sort_by_clustering()
rb$plot()
```

`sort_by_clustering()` options, plus the manual/alphabetical orderings:

```{r sort-options, eval = FALSE}
rb$sort_by_clustering(by_major_allele = TRUE)   # cluster on the major allele only
rb$sort_by_clustering(coverage_cutoff = 0.90)   # only targets in >= 90% of samples
rb$set_sample_order(c("3D7", "HB3"))            # explicit order (rest appended)
rb$sort_alphabetical()                           # back to the default
```

## Plotting: colours

Supply your own ramp (length should match `color_period`):

```{r colors-custom}
rb$plot(colors = c("#F50300", "#FF6E00", "#FFEB01", "#00CA1E", "#0241FE", "#FE00D4"))
```

The package also ships colour-blind-friendly categorical palettes
(`colorPalette_08`, `colorPalette_12`, `colorPalette_15`):

```{r palettes}
scales::show_col(colorPalette_12)
```

The rainbow fill can be driven by either colour column from the prep:

```{r color-col, eval = FALSE}
rb$plot(color_col = "popidFracLogColor")   # default (log-spaced hue)
rb$plot(color_col = "popidFracRegColor")   # linear hue
```

### Discrete rank colours

`plot(rank_colors = TRUE)` colours haplotypes by a discrete rank (with a legend) instead
of the continuous gradient. When there are more haplotype ranks than palette colours, a
colour ramp is interpolated from the palette (dropping white/black and ordering by hue);
if you prepped with `mark_invariant = TRUE`, single-haplotype (invariant) targets get
their own colour (black by default). `get_rank_colors()` returns the rank -> colour map.

```{r rank-colors, fig.width = 14, fig.height = 9}
rb$prep(sort = "population_rank", mark_invariant = TRUE)$sort_by_clustering()
rb$plot(rank_colors = TRUE, x_axis_labels = FALSE)
```

## Plotting: axis labels

Target names (x) and sample names (y) show by default. Toggle either off — the plot
keeps tight margins with no leftover ticks (handy when there are too many samples or
targets to label):

```{r axis-toggle}
rb$plot(x_axis_labels = FALSE)                        # hide target names
```

```{r axis-toggle-both}
rb$plot(x_axis_labels = FALSE, y_axis_labels = FALSE) # hide both
```

## Shade plots

Instead of a rotating rainbow, `prep_shade()` colours each target with shades of a
per-target base colour. `plot()` then uses the shade colours automatically:

```{r shade}
rb$prep_shade(min_pop_size = 1)
rb$plot()
```

`prep_shade()` accepts the base colours to generate shades from:

```{r shade-args, eval = FALSE}
rb$prep_shade(
  min_pop_size = 1,
  bar_height   = 0.80,
  base_colors  = colorPalette_08
)
```

## Interactive tooltips

`plot()` carries the sample, target, haplotype and abundance values as extra aesthetics,
so `plotly::ggplotly()` surfaces them on hover:

```{r plotly, eval = FALSE}
p <- rb$prep(sort = "population_rank")$plot()
plotly::ggplotly(p)
```

## Sample & target metadata

Attach metadata to the object once with `set_sample_meta()` / `set_target_meta()`,
telling it which column matches the sample names / target names. The package ships
example metadata for the isolate data:

```{r set-meta}
data("pfIsosHeomeV1_sampleMeta")   # 61 samples x geography columns
data("pfIsosHeomeV1_targetMeta")   # 100 targets x gene/class columns

rb$prep(sort = "population_rank")$sort_by_clustering()
rb$set_sample_meta(pfIsosHeomeV1_sampleMeta, match_col = "sample")
rb$set_target_meta(pfIsosHeomeV1_targetMeta, match_col = "target_name")
```

`set_*_meta()` **errors** if `match_col` (or a requested `cols`) is absent, and **warns +
NA-fills** any samples/targets that have no metadata row. By default a second call
replaces the stored metadata; pass `add = TRUE` to merge extra columns instead. Restrict
to particular columns with `cols`:

```{r set-meta-cols, eval = FALSE}
rb$set_sample_meta(pfIsosHeomeV1_sampleMeta, match_col = "sample",
                   cols = c("country", "region"))
```

### Sorting by metadata

Order samples (rows) or targets (columns) by metadata; entities with no metadata go to
the end:

```{r sort-meta, eval = FALSE}
rb$sort_samples_by_meta(c("secondaryRegion", "country"))
rb$sort_targets_by_meta("class")
rb$set_target_order(c("Pf01-0145449-0145622"))   # or an explicit target order
```

### The metadata sidebar

`add_sample_metadata()` draws one coloured band per metadata column to the left (or
right) of the rainbow. Sizes are in **cell units**: `width`/`height` of 1 matches a
rainbow cell, 2 is double, 0.8 is 80% (centred); `gap` spaces the bands. Band labels are
drawn on the axis independently of the sample/target names, so you can keep them on even
when the target names are off.

```{r sidebar, fig.width = 16, fig.height = 10}
p <- rb$plot(y_axis_labels = FALSE)
p <- rb$add_sample_metadata(
  p,
  cols          = c("secondaryRegion", "region", "country"),
  width         = 3,     # 3 cells wide
  gap           = 0.5,
  target_labels = FALSE  # hide target names, keep the band labels
)
p
```

Colours are auto-assigned from the colour-blind-friendly palettes; override any column
with a named vector via `colors`. Turn legends off with `legend = FALSE`, or place the
sidebar on the right with `side = "right"`.

### The target annotation strip

`add_target_annotation()` is the transpose: coloured bands above (or below) the rainbow,
one per target-metadata column.

```{r target-strip, fig.width = 16, fig.height = 10}
p <- rb$plot(x_axis_labels = FALSE, y_axis_labels = FALSE)
p <- rb$add_sample_metadata(p, cols = c("region", "country"), width = 3, gap = 0.5,
                            target_labels = FALSE)
p <- rb$add_target_annotation(p, cols = "class", height = 2, sample_labels = FALSE)
p
```

## Saving to PDF, sized automatically

`save_pdf()` sizes the figure from the data (width scales with targets, height with
samples) and uses `cairo_pdf` by default (which keeps hyphens as hyphens). Pass
`device = "pdf"` for the base device (e.g. on Windows). `dims()` returns the computed
size if you want it for a Quarto chunk.

```{r save, eval = FALSE}
rb$dims()                                   # list(width, height) in inches
rb$save_pdf(p, "rainbow.pdf")               # auto-sized, cairo device
rb$save_pdf(p, "rainbow.pdf", device = "pdf", extra_width = 4)  # base device
```

## Accessing the prepped data

`get_prepped()` returns the prepped table if you want to post-process the plot yourself:

```{r get-prepped}
head(rb$get_prepped())
```
