---
title: "Examples: a full tour of the options"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 100,
  out.width = "100%"
)
```

This article is the exhaustive companion to the README: it walks through **every**
`HaplotypeRainbow` option so the README can stay slim. Each section is a self-contained
recipe you can copy.

```{r setup}
library(HaplotypeRainbows)
library(ggplot2)

# example data (uses the older SeekDeep column names)
data("pfisolateExample")
```

## Constructing the object

`HaplotypeRainbow` carries your column mapping so you only set it once. The defaults
follow the Portable Microhaplotype Object (PMO) convention
(`library_sample_name` / `target_name` / `seq` / `reads`); the example data uses the
older SeekDeep names, so we pass them explicitly.

```{r construct}
rb <- HaplotypeRainbow$new(
  pfIsosHeomeV1,
  sample_col    = "s_Sample",
  target_col    = "p_name",
  popuid_col    = "h_popUID",
  rel_abund_col = "c_AveragedFrac"
)
rb
```

`haplotype_rainbow()` is an identical convenience constructor:

```{r construct-wrapper, eval = FALSE}
rb <- haplotype_rainbow(my_pmo_allele_table)   # PMO column names, no mapping needed
```

Transforming methods (`prep()`, `prep_shade()`, `sort_*()`, `set_sample_order()`) mutate
the object and return it, so they chain; `plot()` returns a ggplot.

## Prepping: haplotype ordering

`prep(sort = ...)` controls how haplotypes stack within each cell.

```{r prep-population}
rb$prep(sort = "population_rank")   # order by population rank (default)
rb$plot()
```

```{r prep-frac}
rb$prep(sort = "within_sample_freq")   # order by within-sample fraction
rb$plot()
```

Other `prep()` knobs:

```{r prep-args, eval = FALSE}
rb$prep(
  sort         = "population_rank",
  min_pop_size = 2,     # drop targets with < 2 unique haplotypes
  color_period = 11,    # hue steps to rotate across targets (the "rainbow" period)
  bar_height   = 0.80   # < 1 leaves a gap between sample rows
)
```

## Sorting samples

By default samples are alphabetical. Cluster similar samples (by haplotype sharing,
ward.D2) so they sit together:

```{r sort-cluster}
rb$prep(sort = "population_rank")$sort_by_clustering()
rb$plot()
```

`sort_by_clustering()` options, plus the manual/alphabetical orderings:

```{r sort-options, eval = FALSE}
rb$sort_by_clustering(by_major_allele = TRUE)   # cluster on the major allele only
rb$sort_by_clustering(coverage_cutoff = 0.90)   # only targets in >= 90% of samples
rb$set_sample_order(c("3D7", "HB3"))            # explicit order (rest appended)
rb$sort_alphabetical()                           # back to the default
```

## Plotting: colours

Supply your own ramp (length should match `color_period`):

```{r colors-custom}
rb$plot(colors = c("#F50300", "#FF6E00", "#FFEB01", "#00CA1E", "#0241FE", "#FE00D4"))
```

The package also ships colour-blind-friendly categorical palettes
(`colorPalette_08`, `colorPalette_12`, `colorPalette_15`):

```{r palettes}
scales::show_col(colorPalette_12)
```

The rainbow fill can be driven by either colour column from the prep:

```{r color-col, eval = FALSE}
rb$plot(color_col = "popidFracLogColor")   # default (log-spaced hue)
rb$plot(color_col = "popidFracRegColor")   # linear hue
```

## Plotting: axis labels

Target names (x) and sample names (y) show by default. Toggle either off — the plot
keeps tight margins with no leftover ticks (handy when there are too many samples or
targets to label):

```{r axis-toggle}
rb$plot(x_axis_labels = FALSE)                        # hide target names
```

```{r axis-toggle-both}
rb$plot(x_axis_labels = FALSE, y_axis_labels = FALSE) # hide both
```

## Shade plots

Instead of a rotating rainbow, `prep_shade()` colours each target with shades of a
per-target base colour. `plot()` then uses the shade colours automatically:

```{r shade}
rb$prep_shade(min_pop_size = 1)
rb$plot()
```

`prep_shade()` accepts the base colours to generate shades from:

```{r shade-args, eval = FALSE}
rb$prep_shade(
  min_pop_size = 1,
  bar_height   = 0.80,
  base_colors  = colorPalette_08
)
```

## Interactive tooltips

`plot()` carries the sample, target, haplotype and abundance values as extra aesthetics,
so `plotly::ggplotly()` surfaces them on hover:

```{r plotly, eval = FALSE}
p <- rb$prep(sort = "population_rank")$plot()
plotly::ggplotly(p)
```

## Accessing the prepped data

`get_prepped()` returns the prepped table if you want to post-process the plot yourself:

```{r get-prepped}
rb$prep(sort = "population_rank")
head(rb$get_prepped())
```
