---
title: "Getting started"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 100,
  out.width = "100%"
)
```

HaplotypeRainbows builds haplotype **"rainbow"** plots from targeted-amplicon
(microhaplotype) data: a grid where **columns are targets** and **rows are samples**, and
each cell shows the within-sample haplotype composition. Haplotype colours rotate across
targets to produce the characteristic rainbow.

Everything hangs off one R6 class, `HaplotypeRainbow`, which carries your column mapping
so you only set it once. This article covers the essentials; the other articles go deep
on [colours](colours.html), [metadata](metadata.html),
[clustering & splitting](clustering.html), and [saving & interactivity](saving.html).

```{r setup}
library(HaplotypeRainbows)
library(ggplot2)

# example data (uses the older SeekDeep column names)
data("pfisolateExample")
```

## Constructing the object

You need four columns: a **sample** id, a **target** id, a within-target **haplotype** id,
and a **relative count** (raw counts are fine — they're normalised to within-sample
fractions internally). The defaults follow the Portable Microhaplotype Object (PMO)
convention (`library_sample_name` / `target_name` / `seq` / `reads`); the example data
uses the older SeekDeep names, so we pass them explicitly.

```{r construct}
rb <- HaplotypeRainbow$new(
  pfIsosHeomeV1,
  sample_col    = "s_Sample",
  target_col    = "p_name",
  popuid_col    = "h_popUID",
  rel_abund_col = "c_AveragedFrac"
)
rb
```

`haplotype_rainbow()` is an identical convenience constructor for PMO-named tables:

```{r construct-wrapper, eval = FALSE}
rb <- haplotype_rainbow(my_pmo_allele_table)   # PMO column names, no mapping needed
```

**Chaining.** Transforming methods (`prep()`, `prep_shade()`, `sort_*()`,
`set_sample_order()`, `add_cluster_gaps()`) mutate the object and return it invisibly, so
they chain. `plot()` and the `add_*()` helpers return a ggplot.

```{r chain, eval = FALSE}
rb$prep(sort = "population_rank")$sort_by_clustering()$plot()
```

## Prepping: haplotype ordering

`prep()` must be called before plotting. `sort` controls how haplotypes stack within each
cell:

```{r prep-population}
rb$prep(sort = "population_rank")   # order by population rank (default)
rb$plot()
```

```{r prep-frac}
rb$prep(sort = "within_sample_freq")   # order by within-sample fraction
rb$plot()
```

Other `prep()` knobs — `min_pop_size` drops sparse targets, `color_period` sets the
rainbow period (see the [colours article](colours.html)), and `bar_height < 1` leaves a
gap between sample rows:

```{r prep-barheight}
rb$prep(sort = "population_rank", bar_height = 0.6)   # thinner bars, more row gap
rb$plot(x_axis_labels = FALSE)
```

## Axis labels

Target names (x) and sample names (y) show by default. Toggle either off — the plot keeps
tight margins with no leftover ticks, which is handy when there are too many samples or
targets to label:

```{r axis-toggle}
rb$prep(sort = "population_rank")
rb$plot(x_axis_labels = FALSE)                        # hide target names
```

```{r axis-toggle-both}
rb$plot(x_axis_labels = FALSE, y_axis_labels = FALSE) # hide both
```

## Accessing the prepped data

`get_prepped()` returns the prepped table if you want to post-process the plot yourself or
inspect the derived columns. It uses canonical column names — `sample`, `target`, `hapid`
and `rel_abund` (raw counts), plus `within_sample_freq` / `total_abund` and the derived
geometry/colour columns — regardless of your input column names:

```{r get-prepped}
head(rb$get_prepped())
```

To reconnect the prepped table to your source data, relabel those four columns back to your
original names with `get_prepped(original_names = TRUE)`, or look up the mapping directly
with `column_map()`:

```{r get-prepped-original, eval = FALSE}
rb$column_map()                     # canonical -> your original column names
head(rb$get_prepped(original_names = TRUE))
```
