---
title: "Sample & target metadata"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 16,
  fig.height = 10,
  dpi = 100,
  out.width = "100%"
)
```

Attach per-sample and per-target metadata to the object once, then render it as coloured
**bands** — a sidebar next to the rows, and a strip above/below the columns. This article
covers attaching, sorting by, and drawing metadata, with the full set of sizing and legend
options.

```{r setup}
library(HaplotypeRainbows)
library(ggplot2)

data("pfisolateExample")
data("pfIsosHeomeV1_sampleMeta")   # 61 samples x geography columns
data("pfIsosHeomeV1_targetMeta")   # 100 targets x gene/class columns

rb <- HaplotypeRainbow$new(
  pfIsosHeomeV1,
  sample_col    = "s_Sample",
  target_col    = "p_name",
  popuid_col    = "h_popUID",
  rel_abund_col = "c_AveragedFrac"
)
rb$prep(sort = "population_rank")$sort_by_clustering()
```

## Attaching metadata

`set_sample_meta()` / `set_target_meta()` take the metadata frame and the column that
matches the sample / target names:

```{r set-meta}
rb$set_sample_meta(pfIsosHeomeV1_sampleMeta, match_col = "sample")
rb$set_target_meta(pfIsosHeomeV1_targetMeta, match_col = "target_name")
```

They **error** if `match_col` (or a requested `cols`) is absent, and **warn + NA-fill** any
samples/targets that have no metadata row. A second call replaces the stored metadata by
default; pass `add = TRUE` to merge extra columns. Restrict to particular columns with
`cols`:

```{r set-meta-cols, eval = FALSE}
rb$set_sample_meta(pfIsosHeomeV1_sampleMeta, match_col = "sample",
                   cols = c("country", "region"))
rb$set_sample_meta(more_meta, match_col = "sample", add = TRUE)   # merge, don't replace
```

### Sorting by metadata

Order rows or columns by metadata; entities with no metadata go to the end:

```{r sort-meta, eval = FALSE}
rb$sort_samples_by_meta(c("secondaryRegion", "country"))
rb$sort_targets_by_meta("class")
rb$set_target_order(c("Pf01-0145449-0145622"))   # or an explicit target order
```

## The sample sidebar

`add_sample_metadata()` draws one coloured band per metadata column to the left (or right)
of the rainbow. Sizes are in **cell units**: `width`/`height` of 1 matches a rainbow cell,
2 is double, 0.8 is 80% (centred). Band labels are drawn independently of the sample /
target names, so you can keep them on when the target names are off:

```{r sidebar}
p <- rb$plot(y_axis_labels = FALSE)
rb$add_sample_metadata(
  p,
  cols          = c("secondaryRegion", "region", "country"),
  width         = 3,     # 3 cells wide
  gap           = 0.5,   # gap between bands
  target_labels = FALSE  # hide target names, keep the band labels
)
```

Colours are auto-assigned from the colour-blind palettes. The sidebar has knobs for
placement, spacing, and — importantly for wide metadata — legend layout and order:

- `side = "right"` moves it to the other side.
- `plot_gap` separates the bands from the cells (default 1).
- `legend_ncol` / `legend_nrow` reshape big legends (a scalar for all bands, or a named
  vector per column) so they stay on the page.
- `level_order` (a named list per column) controls the legend entry order; unlisted values
  are appended in sorted order.
- `colors` overrides the auto colours per column; `na_color` sets the missing-value fill;
  `legend = FALSE` drops the legends.

```{r sidebar-tweaks}
p <- rb$plot(y_axis_labels = FALSE)
rb$add_sample_metadata(
  p,
  cols        = c("region", "country"),
  side        = "right",
  plot_gap    = 2,                                     # push off the rainbow
  legend_ncol = c(country = 4, region = 2),            # fold tall legends
  level_order = list(region = c("West Africa", "East Africa")),
  target_labels = FALSE
)
```

## The target strip

`add_target_annotation()` is the transpose: coloured bands above (`position = "top"`, the
default) or below the rainbow, one per target-metadata column. Stack multiple columns and
space them with `gap`:

```{r target-strip}
p <- rb$plot(x_axis_labels = FALSE, y_axis_labels = FALSE)
p <- rb$add_sample_metadata(p, cols = c("region", "country"), width = 3, gap = 0.5,
                            target_labels = FALSE)
# two stacked target layers: class (innermost) then chrom
rb$add_target_annotation(p, cols = c("class", "chrom"), height = 2, gap = 0.5,
                         plot_gap = 1, sample_labels = FALSE)
```

Put the strip below instead, and reshape its legend the same way as the sidebar:

```{r target-strip-bottom, eval = FALSE}
rb$add_target_annotation(p, cols = "chrom", position = "bottom",
                         legend_ncol = 2, level_order = list(chrom = c("Pf3D7_01_v3")))
```
